A Decision Procedure for (Co)datatypes in SMT Solvers

### 2016-02-09 sister/sister.tex
more

\section{Implementation and Evaluation} % on Isabelle Problems}
\label{sec:evaluation}

\newcommand\gandl{SBT}
\newcommand\HD[1]{\hbox to2em{\footnotesize\hfill\!\!\!\!#1\!\!\!\!\hfill}}

@@

By contrast, the contributions of the decision procedure
to proving are less obvious; they depend on how
often acyclicity and uniqueness are necessary for a proof.
To evaluate this,
we generated benchmark problems from existing interactive
proof goals arising in existing
Isabelle formalizations, using Sledgehammer \cite{blanchette-et-al-2013-smt}
as translator.
%There are not yet many benchmarks available, since codatatypes are a
%recent addition to Isabelle \cite{blanchette-et-al-2014-impl}. \rem{Necessary to mention this? Seems to detract from significance of contribution.}
We
included all the formalizations from the Isabelle distribution (Distro, 1179~goals)
and the \emph{Archive of Formal Proofs} (AFP, 3014~goals) that define codatatypes falling within the supported fragment. We
also included formalizations about Bird and Stern--Brocot trees (\gandl,
265~goals). % \cite{gammie-lochbihler-2016}.
To exercise the datatype support, formalizations about finite
lists and trees
were added to the first two benchmark sets. The formalizations were selected before
conducting any experiments.
%The experimental data are available
%online.%
%\footnote{\url{http://lara.epfl.ch/~reynolds/JAR-CADE2015-cdt/}}

@@

For each proof goal in each formalization, we used Sledgehammer to select either 16 or
256~lemmas, which were monomorphized and translated to SMT-LIB along with the
goal. The resulting problem was given to the
development version of CVC4
(from 15 September 2015) and to
Z3 4.3.2 for comparison, each running for
up to 60~s on StarExec \cite{stump-et-al-2014-starexec}.
Problems not involving any (co)datatypes were left out.
Due to the lack of machinery in Isabelle for parsing CVC4 proofs and
reconstructing inferences about (co)datatypes, the solvers are trusted
as oracles. % in these experiments.
%Z3 is included for comparison.

CVC4 was run on each problem several times, with the
support for datatypes and codatatypes either enabled or disabled. The
contributions of the acyclicity and uniqueness rules were also measured,
by selectively enabling or disabling the rules.
% to find
% out whether these somewhat expensive rules are useful in practice.
Even when the
decision procedure is disabled, the problems may contain
basic lemmas about constructors and selectors, allowing some (co)datatype
reasoning. This is especially true for problems generated using 256~lemmas.
The problems with 16~lemmas put more stress on the decision procedure but are
less typical of Sledgehammer-generated problems.

@@

\begin{table*}[t!]
\normalsize
\begin{center}\begin{tabular}{@{\;}l@{\kern1.5em}c@{\kern.5em}c@{\kern1em}c@{\kern.5em}c@{\kern1em}c@{\kern.5em}c@{\kern1.5em}c@{\kern.5em}c@{\;}}
  & \multicolumn{2}{@{}c@{\kern1em}}{Distro} & \multicolumn{2}{@{}c@{\kern1em}}{AFP} & \multicolumn{2}{@{}c@{\kern1.5em}}{\gandl} & \multicolumn{2}{@{}c@{\;}}{Overall} \\[-0.5pt]
  & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{Z3} & \HD{CVC4}& \HD{Z3} 
\MIDRULE
No (co)datatypes
  & 287 & 282 & 765 & 771 & 47 & 44 & 1099 & 1097 \\%[\jot]
%Datatype simplification only
%  &  \\
Datatypes without \rn{Acyclic}
  & 298 & -- & 771 & -- & 47 & -- & 1116 & -- \\
Full datatypes
  & 298 & 294 & 775 & 783 & 47 & 44 & 1120 & 1121 \\%[\jot]
%Codatatype simplification only
%  &  \\
Codatatypes without \rn{Unique}\!\!
  & 288 & -- & 797 & -- & 47 & -- & 1132 & -- \\
Full codatatypes
  & 288 & -- & 797 & -- & \bfseries 52 & -- & 1137 & -- \\%[\jot]
Full (co)datatypes
  & \bfseries 299 & -- & \bfseries 806 & -- & \bfseries 52 & -- & \bfseries 1157 & -- %\\[\jot]
\end{tabular}\end{center}
\caption{\captioni}
\label{tab:benchsixteen}
\end{table*}

\begin{table*}[t!]
\normalsize
\begin{center}\begin{tabular}{@{\;}l@{\kern1.5em}c@{\kern.5em}c@{\kern1em}c@{\kern.5em}c@{\kern1em}c@{\kern.5em}c@{\kern1.5em}c@{\kern.5em}c@{\;}}
  & \multicolumn{2}{@{}c@{\kern1em}}{Distro} & \multicolumn{2}{@{}c@{\kern1em}}{AFP} & \multicolumn{2}{@{}c@{\kern1.5em}}{\gandl} & \multicolumn{2}{@{}c@{\;}}{Overall} \\[-0.5pt]
  & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{Z3} & \HD{CVC4}& \HD{Z3} 
\MIDRULE
No (co)datatypes
  & 617 & 560 & 1503 & 1271 & 89 & 80 & 2209 & 1911 \\%[\jot]
%Datatype simplification only
%  &  \\
Datatypes without \rn{Acyclic}
  & 617 & -- & \bfseries 1504 & -- & 90 & -- & 2211 & -- \\
Full datatypes
  & 617 & 560 & \bfseries 1504 & 1263 & 90 & 78 & 2211 & 1901 \\%[\jot]
%Codatatype simplification only
%  &  \\
Codatatypes without \rn{Unique}\!\!
  & 617 & -- & 1501 & -- & 90 & -- & 2208 & -- \\
Full codatatypes
  & \bfseries 620 & -- & 1502 & -- & 98 & -- & \bfseries 2220 & -- \\%[\jot]
Full (co)datatypes
  & 619 & -- & 1501 & -- & \bfseries 99 & -- & 2219 & -- %\\[\jot]
\end{tabular}\end{center}
\caption{\captionii}
\label{tab:benchtwofiftysix}
\end{table*}

The results are summarized in Tables \ref{tab:benchsixteen} and
\ref{tab:benchtwofiftysix}.
%The goals in Table \ref{tab:benchsixteen} include 16
%heuristically selected lemmas (before monomorphization);
%\cite{meng-paulson-2009-relev};
%that number is increased to 256 for Table \ref{tab:benchtwofiftysix}.
The decision procedure
makes a difference across all three benchmark suites. For the 16-lemma
problems, it accounts for an overall
success rate increase of over 5\%.
Moreover, every aspect of the procedure, including the
more expensive rules, makes a contribution.
%
For the 256-lemma problems, the difference
is much smaller, at 0.5\%.
%For Isabelle
%users, the practical usefulness of interpreted (co)datatypes is roughly
%comparable to that of arithmetic \cite{blanchette-et-al-2013-smt}, although this
%obviously depends on the application.
Table~\ref{tab:benchtwofiftysix} indicates that
the theoretically stronger instances of the decision
procedure do not always subsume the weaker ones in practice.
The raw data reveal that the full procedure proved 27~goals that
could not be proved without it, but failed for 17~goals that could be proved
without it. This potentially points to poor interactions between the decision
procedure and the quantifier instantiation module
\cite{ReynoldsTinelliMoura14}.

>>>>

\section{Evaluation} % on Isabelle Problems}
\label{sec:evaluation}

\newcommand\gandl{SBT}
\newcommand\HD[1]{\hbox to2em{\relax\hfill\!\!\!\!#1\!\!\!\!\hfill}}

@@

By contrast, the contributions of the decision procedure
to proving are less obvious; they depend on how
often acyclicity and uniqueness are necessary for a proof.
To evaluate this,
we generated benchmark problems from existing interactive
proof goals arising in existing
Isabelle formalizations, using Sledgehammer \cite{blanchette-et-al-2013-smt}
as translator.
%There are not yet many benchmarks available, since codatatypes are a
%recent addition to Isabelle \cite{blanchette-et-al-2014-impl}. \rem{Necessary to mention this? Seems to detract from significance of contribution.}
We
included all the formalizations from the Isabelle distribution (Distro, 1179~goals)
and the \emph{Archive of Formal Proofs} (AFP, 3014~goals) that define codatatypes falling within the supported fragment. We
also included formalizations about Bird and Stern--Brocot trees (\gandl,
265~goals). % \cite{gammie-lochbihler-2016}.
To exercise the datatype support, formalizations about finite
lists and trees
were added to the first two benchmark sets.
%The formalizations were selected before
%conducting any experiments.
%The experimental data are available
%online.%
%\footnote{\url{http://lara.epfl.ch/~reynolds/JAR-CADE2015-cdt/}}

@@

For each proof goal in each formalization, we used Sledgehammer to select either 16 or
256~lemmas, which were monomorphized and translated to SMT-LIB along with the
goal. The resulting problem was given to the
development version of CVC4
(from 15 September 2015) and to
Z3 4.3.2 for comparison, each running for
up to 60~s. % on StarExec \cite{stump-et-al-2014-starexec}.
Problems not involving any (co)datatypes were left out.

CVC4 was run on each problem several times, with the
support for datatypes and codatatypes either enabled or disabled. The
contributions of the acyclicity and uniqueness rules were also measured,
by selectively enabling or disabling the rules.
% to find
% out whether these somewhat expensive rules are useful in practice.
Even when the
decision procedure is disabled, the problems may contain
basic lemmas about constructors and selectors, allowing some (co)datatype
reasoning. This is especially true for problems generated using 256~lemmas.
The problems with 16~lemmas put more stress on the decision procedure but are
less typical of Sledgehammer problems.

@@

%%% TYPESETTING: hack
\setcounter{table}{3}

\begin{table}[t!]
\normalsize
\begin{center}\begin{tabular}{@{\;}l@{\kern1.5em}c@{\kern.5em}c@{\kern1em}c@{\kern.5em}c@{\;}}
  & \multicolumn{2}{@{}c@{\kern1em}}{16 lemmas\;} & \multicolumn{2}{@{}c@{\;}}{256 lemmas} \\%[-0.5pt]
  & \HD{CVC4} & \HD{Z3} & \HD{CVC4} & \HD{\,Z3}
\MIDRULE
No (co)datatypes
 &  1099  &  1097  &  2209  &  1911 \\%[\jot]
Datatypes without \rn{Acyclic}
 &  1116  &  -- &  2211  &  -- \\
Full datatypes
 &  1120  &  1121  &  2211  &  1901 \\%[\jot]
Codatatypes without \rn{Unique}\!\!
 &  1132  &  --  &  2208  &  -- \\
Full codatatypes
 &  1137  &  -- &  \bfseries 2220  &  -- \\%[\jot]
Full (co)datatypes
 &  \bfseries 1157  &  -- &  2219  &  -- %\\[\jot]
\end{tabular}\end{center}
\caption{\captioni}
\label{tab:bench}
\end{table}

The results are summarized in Table \ref{tab:bench}.
For the 16-lemma problems, it accounts for an overall
success rate increase of over 5\%.
Moreover, every aspect of the procedure, including the
more expensive rules, makes a contribution.
%
For the 256-lemma problems, the difference
is much smaller, at 0.5\%.
%For Isabelle
%users, the practical usefulness of interpreted (co)datatypes is roughly
%comparable to that of arithmetic \cite{blanchette-et-al-2013-smt}, although this
%obviously depends on the application.
Table~\ref{tab:benchtwofiftysix} indicates that
the theoretically stronger instances of the decision
procedure do not always subsume the weaker ones in practice.
The raw data reveal that the full procedure proved 27~goals that
could not be proved without it, but failed for 17~goals that could be proved
without it.
%This potentially points to poor interactions between the decision
%procedure and the quantifier instantiation module
%\cite{ReynoldsTinelliMoura14}.
